Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-04-14 paper.tex
Adding references

The Blue Obelisk was first described in Guha et al.\cite{guha2006}. This
contribution looks back on the past 5 years and surveys progress and
remaining challenges in the areas of Open Source, Open Data and Open
Standards in chemistry.

@@

Bioclipse (v1.0 released in Aug 2006 - ????within the scope of the paper???) and Open Babel (v1.0 in Oct 2009) are two examples of such software, based on the CDK and Open Babel, respectively. Bioclipse is an award-winning `molecular workbench' for life sciences that ..., while Avogadro is a 3D molecular editor and viewer aimed at preparing and analysis computational chemistry calculations. An interesting aspect of both of these projects is that they share some developers with the underlying toolkit and this has driven the development of new features in the CDK and Open Babel (??? is this true for the CDK too? ???).

@@

Although the software developed by the Blue Obelisk is Open Source on principle, this does not rule out the establishment of a business based on open source chemistry software. Rather than supporting itself through sales, such a business generally relies on providing support, customisation and services around the software. The interaction between an open source project and a company is important, and works best where there is some form of ``giving back" to the project (such as developer time, code contributions, or finanical assistance with server costs).

@@

One of the goals of the Blue Obelisk is to promote interoperability in chemical
informatics. When barriers exist to moving chemical data between
different software, the community becomes fragmented and there is
the danger of vendor lock-in (where users are constrained to using
a particular software, a situation which puts them at a
disadvantage). This applies as much to Open Source software as to
proprietary software. Cinfony is a project (first release May 2008)
whose goal is to tackle this problem in the area of cheminformatics toolkits.
It is a Python library that enables Open Babel, the CDK, and RDKit to
be used using the same API; this makes it easy, for example, to read a
molecule using Open Babel, calculate descriptors using the CDK and
create a depiction using RDKit.

@@

A core requirement for chemical structure databases and chemical
registration systems in general is the notion of structure
standardization.  That is,  for a given input structure, multiple
representations should be converted to one canonical form. 
Structure canonicalization routines partially address this aspect,
converting multiple alternative topologies to a single canonical
form. However, the problem of standardization is broader than just
topological canonicalization. Features that must be considered include
\begin{itemize}
\item topological canonicalization
\item handling of charges
\item tautomer enumeration and canonicalization
\item normalization of functional groups
\end{itemize}
Currently, most of the individual components of a ``standardization
pipeline'' can be implemented using BO tools. The larger problem is
that there is no agreed upon list of steps for a standardization
process. While some specifications have been published (e.g., Pubchem)
and some standardization services and tools are available (Pubchem
provides an online service to standaridize molecules and the NCGC
provides a stand alone tool) each group has their own set of rules. A
common reference specification for standardization would be of immense
value in interoperability between structure repositories as well as
between toolkits (though the latter is still confounded by differences
in lower level cheminformatic features such as aromaticity models).

@@

A considerable stumbling block in advocating the release of scientific
data as Open Data has been how exactly to define Open. A major step
forward was the launch in 2010 of the Panton Principles for Open Data
in Science. This formalises the idea that Open Data maximises the
possibility of reuse and repurposing, the fundamental basis
of how science works. These principles recommend that published data
be licensed explicitly, and preferably under CC0 (Creative Commons "No
Rights Reserved"). This license allows others to use the data for any
purpose whatsoever without any barriers.

@@

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
 \bibliographystyle{bmc_article}  % Style BST file
  \bibliography{paper} }     % Bibliography file (usually '*.bib' )

>>>>

The Blue Obelisk was first described in Guha et al.\cite{Guha2006} in
2006. This
contribution looks back on the past 5 years and surveys progress and
remaining challenges in the areas of Open Source, Open Data and Open
Standards in chemistry.

@@

Bioclipse \cite{Bioclipse2} (v1.0 released in Aug 2006) and Avogadro
\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based on the CDK and Open Babel, respectively. Bioclipse is an award-winning `molecular workbench' for life sciences that ..., while Avogadro is a 3D molecular editor and viewer aimed at preparing and analysis computational chemistry calculations. An interesting aspect of both of these projects is that they share some developers with the underlying toolkit and this has driven the development of new features in the CDK and Open Babel (??? is this true for the CDK too? ???).

@@

Although the software developed by the Blue Obelisk is Open Source on principle, this does not rule out the establishment of a business based on open source chemistry software. Rather than supporting itself through sales, such a business generally relies on providing support, customisation and services around the software. The interaction between an open source project and a company is important, and works best where there is some form of `giving back' to the project (such as developer time, code contributions, or finanical assistance with server costs).

@@

One of the goals of the Blue Obelisk is to promote interoperability in chemical
informatics. When barriers exist to moving chemical data between
different software, the community becomes fragmented and there is
the danger of vendor lock-in (where users are constrained to using
a particular software, a situation which puts them at a
disadvantage). This applies as much to Open Source software as to
proprietary software. Cinfony is a project (first release in May 2008)
whose goal is to tackle this problem in the area of cheminformatics
toolkits \cite{OBoyleCinfony2008}.
It is a Python library that enables Open Babel, the CDK, and RDKit to
be used using the same API; this makes it easy, for example, to read a
molecule using Open Babel, calculate descriptors using the CDK and
create a depiction using RDKit.

@@

A core requirement for chemical structure databases and chemical
registration systems in general is the notion of structure
standardization.  That is,  for a given input structure, multiple
representations should be converted to one canonical form. 
Structure canonicalization routines partially address this aspect,
converting multiple alternative topologies to a single canonical
form. However, the problem of standardization is broader than just
topological canonicalization. Features that must be considered include
\begin{itemize}
\item topological canonicalization
\item handling of charges
\item tautomer enumeration and canonicalization
\item normalization of functional groups
\end{itemize}
Currently, most of the individual components of a `standardization
pipeline' can be implemented using BO tools. The larger problem is
that there is no agreed upon list of steps for a standardization
process. While some specifications have been published (e.g., Pubchem)
and some standardization services and tools are available (Pubchem
provides an online service to standaridize molecules and the NCGC
provides a stand alone tool) each group has their own set of rules. A
common reference specification for standardization would be of immense
value in interoperability between structure repositories as well as
between toolkits (though the latter is still confounded by differences
in lower level cheminformatic features such as aromaticity models).

@@

A considerable stumbling block in advocating the release of scientific
data as Open Data has been how exactly to define Open. A major step
forward was the launch in 2010 of the Panton Principles for Open Data
in Science \cite{WebPanton}. This formalises the idea that Open Data maximises the
possibility of reuse and repurposing, the fundamental basis
of how science works. These principles recommend that published data
be licensed explicitly, and preferably under CC0 (Creative Commons `No
Rights Reserved') \cite{WebCC0}. This license allows others to use the data for any
purpose whatsoever without any barriers.

@@

{\ifthenelse{\boolean{publ}}{\footnotesize}{\small}
 \bibliographystyle{bmc_article}  % Style BST file
  \bibliography{websites,paper} }     % Bibliography file (usually '*.bib' )
